## Hands-on worksheet

> You will do this inside your Codespace, in `notebooks/01-visualization.qmd`.
> The same steps are shown here so you can read along.

### The one big idea

Every **row** in a table becomes one **mark** on a plot. The **geom** decides
what kind of mark (dots, boxes, bars) and how to draw it. The **mapping**,
written inside `aes()`, links a **column** to a **property** of the mark (its
x position, its y position, its colour).

### Read a plot

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(palmerpenguins)

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point()
```

- **Mark:** a dot. **Mapping:** `flipper_length_mm` → x, `body_mass_g` → y,
  `species` → colour. **Rows per mark:** one row = one dot.

### Build it up, one line at a time

Start simple, then add one piece at a time:

```{r}
#| eval: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()                          # step 1: bare scatter
```

Add `color = species` **inside** `aes()` to map colour to a column. Compare that
with putting `color = "navy"` **outside** `aes()` (inside `geom_point()`), which
sets every dot to the same colour — a setting, not a mapping.

### Change the geom, keep the rules

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot() +
  labs(x = "Species", y = "Body mass (g)", fill = "Species")
```

Same idea, new mark: each **box** summarises **many** rows (a five-number
summary) instead of one row per mark.

### One wrangling move: the pipe

`|>` means "take this data, then…". Filter first, then plot:

```{r}
penguins |>
  filter(species == "Gentoo") |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```
