## Hands-on worksheet

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(palmerpenguins)
```

> You'll do the real work in your Codespace, in `notebooks/01-visualization.qmd`.
> This page lists the **tasks**. Try each one **yourself first** — the
> collapsible **"Show an example"** boxes are hints to peek at only if you get
> stuck.

Each task has a number so we can point to it out loud. The first number is the
exercise and the second is the task — so **2.3** means "Exercise 2, task 3."

## Exercise 1 — The elements of a plot

Every **row** in a table becomes one **mark** on a plot. The **geom** decides
what *kind* of mark; the **mapping** (written inside `aes()`) links a **column**
to a property of the mark. We use the `palmerpenguins` data.

**1.1 — Read a plot.** Look at the plot below and answer in your own words:
(a) What is the **mark**? (b) Which column controls **x**, **y**, and **colour**?
(c) How many **rows** of the table made one mark?

```{r}
#| echo: false
ggplot(penguins, aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point()
```

::: {.callout-tip collapse="true" title="Show the answer"}
- **Mark:** a dot. **Mappings:** `flipper_length_mm` → x, `body_mass_g` → y,
  `species` → colour. **Rows per mark:** one row = one dot.
:::

**1.2 — Build a scatter.** Make a scatter plot of `flipper_length_mm` (x)
against `body_mass_g` (y) using `geom_point()`.

::: {.callout-tip collapse="true" title="Show an example of a working plot"}
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```
:::

**1.3 — Colour by a column.** Add `color = species` **inside** `aes()` so each
species gets its own colour.

::: {.callout-tip collapse="true" title="Show an example of a working plot"}
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point()
```
:::

**1.4 — Mapping vs. setting.** Make *all* the dots the same fixed colour (say,
navy). Does `color = "navy"` go **inside** `aes()` or **outside**? Why?

::: {.callout-tip collapse="true" title="Show the answer"}
It goes **outside** `aes()`, inside the geom — it's a fixed value, not a column
mapping:
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(color = "navy")
```
:::

**1.5 — Change the geom.** Compare body mass across species with a boxplot: map
`species` to x, `body_mass_g` to y, and use `geom_boxplot()`.

::: {.callout-tip collapse="true" title="Show an example of a working plot"}
```{r}
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot()
```
:::

**1.6 — Filter with the pipe.** Use `|>` to keep only one species (your choice),
then make a scatter of just those penguins.

::: {.callout-tip collapse="true" title="Show an example of a working plot"}
```{r}
penguins |>
  filter(species == "Gentoo") |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```
:::

## Exercise 2 — The shape of one variable

A **histogram** cuts one numeric column into bins and counts the rows in each. A
**boxplot** is a compact summary, handy for comparing groups.

**2.1 — Make a histogram** of `flipper_length_mm`.

::: {.callout-tip collapse="true" title="Show an example of a working plot"}
```{r}
ggplot(penguins, aes(x = flipper_length_mm)) +
  geom_histogram(binwidth = 5)
```
:::

**2.2 — Change the binwidth.** Try `binwidth = 2`, then `binwidth = 20`. In a
sentence, what happens to the shape? (No new plot to hand in — just look.)

**2.3 — Compare groups with a boxplot** of `body_mass_g` for each `species`.

::: {.callout-tip collapse="true" title="Show an example of a working plot"}
```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```
:::

## Exercise 3 — Tidy data: reshape, then plot

Real data often comes off a machine in **wide** format — one row per sample, with
a separate column for each measurement. Here is a small plate-reader result for
four genes in ten samples:

```{r}
#| echo: false
plate <- tribble(
  ~sample, ~group,    ~gene_A, ~gene_B, ~gene_C, ~gene_D,
  "s01",   "control",     102,      88,      41,      59,
  "s02",   "control",      98,      91,      44,      62,
  "s03",   "control",     105,      86,      39,      58,
  "s04",   "control",     100,      93,      43,      61,
  "s05",   "control",      97,      89,      40,      60,
  "s06",   "treated",     101,      90,      87,     108,
  "s07",   "treated",      99,      92,      91,     112,
  "s08",   "treated",     104,      87,      85,     105,
  "s09",   "treated",      98,      94,      90,     110,
  "s10",   "treated",     103,      88,      88,     107
)
plate
```

> In your Codespace you'll read this from a file with
> `read_csv("data/qpcr_plate_wide.csv")` — same data, straight off the machine.

**3.1 — Read the table.** What does one **row** represent, and how many separate
measurements are packed into each row?

::: {.callout-tip collapse="true" title="Show the answer"}
Each row is one **sample**, and it holds **four** measurements
(`gene_A`–`gene_D`). That is *wide* format — `ggplot` can't plot it directly yet.
:::

**3.2 — Reshape to long.** Use `pivot_longer()` to stack the four gene columns
into a `gene` column and a `signal` column.

::: {.callout-tip collapse="true" title="Show an example"}
```{r}
plate_long <- plate |>
  pivot_longer(gene_A:gene_D, names_to = "gene", values_to = "signal")
plate_long
```
:::

**3.3 — Plot it.** Make a boxplot of `signal` by `gene`, filled by `group`.
Which genes change between the `control` and `treated` samples?

::: {.callout-tip collapse="true" title="Show an example of a working plot"}
```{r}
ggplot(plate_long, aes(x = gene, y = signal, fill = group)) +
  geom_boxplot()
```
:::
