## Hands-on worksheet

> You will do this inside your Codespace, in `notebooks/01-visualization.qmd`.
> The same steps are shown here so you can read along. We work in three short
> parts, with a little teaching before each.

### The one big idea

Every **row** in a table becomes one **mark** on a plot. The **geom** decides
what kind of mark (dots, boxes, bars) and how to draw it. The **mapping**,
written inside `aes()`, links a **column** to a **property** of the mark (its
x position, its y position, its colour).

## Part A — Read and build a plot

### Read a plot

```{r}
#| warning: false
#| message: false
library(tidyverse)
library(palmerpenguins)

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point()
```

- **Mark:** a dot. **Mapping:** `flipper_length_mm` → x, `body_mass_g` → y,
  `species` → colour. **Rows per mark:** one row = one dot.

### Build it up, one line at a time

Start simple, then add one piece at a time:

```{r}
#| eval: false
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()                          # step 1: bare scatter
```

Add `color = species` **inside** `aes()` to map colour to a column. Compare that
with putting `color = "navy"` **outside** `aes()` (inside `geom_point()`), which
sets every dot to the same colour — a setting, not a mapping.

### Change the geom, keep the rules

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot() +
  labs(x = "Species", y = "Body mass (g)", fill = "Species")
```

Same idea, new mark: each **box** summarises **many** rows (a five-number
summary) instead of one row per mark.

### One wrangling move: the pipe

`|>` means "take this data, then…". Filter first, then plot:

```{r}
penguins |>
  filter(species == "Gentoo") |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

## Part B — The shape of one variable

Sometimes you have a single column of numbers and want to see its **shape**.
A **histogram** cuts the range into bins and counts how many rows land in each:

```{r}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250)
```

Try changing `binwidth` (e.g. `100`, `500`) and watch the shape change.

A **boxplot** squeezes each distribution into one box, which is handy for
comparing groups:

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

Both answer the same question — *what do these numbers look like?* — a histogram
in full detail, a boxplot as a compact summary for comparing.

## Part C — Tidy data: reshape, then plot

Real data often comes off a machine in **wide** format — one row per sample, with
a separate column for each measurement. Here is a small plate-reader result for
four genes measured in ten samples:

```{r}
plate <- tribble(
  ~sample, ~group,    ~gene_A, ~gene_B, ~gene_C, ~gene_D,
  "s01",   "control",     102,      88,      41,      59,
  "s02",   "control",      98,      91,      44,      62,
  "s03",   "control",     105,      86,      39,      58,
  "s04",   "control",     100,      93,      43,      61,
  "s05",   "control",      97,      89,      40,      60,
  "s06",   "treated",     101,      90,      87,     108,
  "s07",   "treated",      99,      92,      91,     112,
  "s08",   "treated",     104,      87,      85,     105,
  "s09",   "treated",      98,      94,      90,     110,
  "s10",   "treated",     103,      88,      88,     107
)
plate
```

> In your Codespace you will instead **read this from a file** with
> `read_csv("data/qpcr_plate_wide.csv")` — same data, straight off the machine.

ggplot wants **tidy** data: one row per observation, one column per variable. One
gene measured in one sample is an observation, so it should be its own row. Use
`pivot_longer()` to stack the four gene columns into two — a `gene` column and a
`signal` column:

```{r}
plate_long <- plate |>
  pivot_longer(gene_A:gene_D, names_to = "gene", values_to = "signal")
plate_long
```

Now every row is one measurement, and we can plot it — comparing the two groups
gene by gene:

```{r}
ggplot(plate_long, aes(x = gene, y = signal, fill = group)) +
  geom_boxplot()
```

The wide table could not make this plot. One `pivot_longer()` unlocked it — and
the pattern (genes C and D switch on with treatment) jumps right out.
