---
title: "Intro to Data Visualization in R"
subtitle: "<em>RISE 2026</em>"
format: qmdrslides-revealjs
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
library(tidyverse)
library(palmerpenguins)
theme_set(theme_minimal(base_size = 18))
```

## Why make a plot?

Four datasets. Same summary statistics. Very different stories.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.5
datasets::anscombe |>
  pivot_longer(everything(),
    names_to = c(".value", "set"),
    names_pattern = "(.)(.)") |>
  ggplot(aes(x, y)) +
  geom_point(size = 2, color = "#0072B2") +
  facet_wrap(~ set, nrow = 1) +
  labs(x = NULL, y = NULL)
```

A picture shows what a table of numbers hides.

## The one big idea

**Every row in your table becomes one mark on the plot.**

::: {.fragment}
- **geom** — the *kind* of mark and the rule for drawing it
- **mark** — what you actually see
- **mapping** — links a *column* to a *property* of the mark (in `aes()`)
:::

## Three words, one plot

```{r}
#| echo: true
#| fig-width: 7
#| fig-height: 4
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point()
```

Mark = dot · Mapping = flipper→x, mass→y, species→colour · one row = one dot.

## Anatomy of `ggplot()`

```r
ggplot(penguins,                    # the data
       aes(x = flipper_length_mm,   # mapping: column -> x
           y = body_mass_g,         # mapping: column -> y
           color = species)) +      # mapping: column -> colour
  geom_point()                      # geom: draw a dot per row
```

## Same data, different geom

```{r}
#| echo: true
#| fig-width: 7
#| fig-height: 3.8
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot()
```

A **box** summarises **many** rows. The mappings still mean the same thing.

## Mapping vs. setting

::: {.columns}
::: {.column width="50%"}
**Inside `aes()` = mapping** (from data)
```r
geom_point(aes(color = species))
```
:::
::: {.column width="50%"}
**Outside `aes()` = setting** (one fixed value)
```r
geom_point(color = "navy")
```
:::
:::

The most common beginner mix-up. Colour from a **column** vs. one **fixed** colour.

## The pipe: `|>`

`|>` means **"take this data, then…"**

```{r}
#| echo: true
#| fig-width: 6
#| fig-height: 3.4
penguins |>
  filter(species == "Gentoo") |>
  ggplot(aes(flipper_length_mm, body_mass_g)) +
  geom_point()
```

## Your turn

Open `notebooks/01-visualization.qmd` and, for each plot, ask:

1. What is the **mark**?
2. Which **column** → which **property**?
3. How many **rows** make one mark?
