---
title: "Intro to Data Visualization in R"
subtitle: "<em>RISE 2026</em>"
format: qmdrslides-revealjs
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| echo: false
library(tidyverse)
library(palmerpenguins)
theme_set(theme_minimal(base_size = 18))

# Small wide "plate reader" dataset used in the Tidy data section.
# One row per sample; each gene column is a separate measurement.
plate <- tribble(
  ~sample, ~group,    ~gene_A, ~gene_B, ~gene_C, ~gene_D,
  "s01",   "control",     102,      88,      41,      59,
  "s02",   "control",      98,      91,      44,      62,
  "s03",   "control",     105,      86,      39,      58,
  "s04",   "control",     100,      93,      43,      61,
  "s05",   "control",      97,      89,      40,      60,
  "s06",   "treated",     101,      90,      87,     108,
  "s07",   "treated",      99,      92,      91,     112,
  "s08",   "treated",     104,      87,      85,     105,
  "s09",   "treated",      98,      94,      90,     110,
  "s10",   "treated",     103,      88,      88,     107
)

plate_long <- plate |>
  pivot_longer(gene_A:gene_D, names_to = "gene", values_to = "signal")
```

## Let's warm up {.your-turn}

{{< include _wooclap.qmd >}}

## Why make a plot?

Four datasets. Same summary statistics. Very different stories.

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 3.5
datasets::anscombe |>
  pivot_longer(everything(),
    names_to = c(".value", "set"),
    names_pattern = "(.)(.)") |>
  ggplot(aes(x, y)) +
  geom_point(size = 2, color = "#0072B2") +
  facet_wrap(~ set, nrow = 1) +
  labs(x = NULL, y = NULL)
```

A picture shows what a table of numbers hides.

## The one big idea

**Every row in your table becomes one mark on the plot.**

::: {.fragment}
- **geom** — the *kind* of mark and the rule for drawing it
- **mark** — what you actually see
- **mapping** — links a *column* to a *property* of the mark (in `aes()`)
:::

## Three words, one plot

```{r}
#| echo: true
#| fig-width: 7
#| fig-height: 4
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point()
```

Mark = dot · Mapping = flipper→x, mass→y, species→colour · one row = one dot.

## Anatomy of `ggplot()`

```r
ggplot(penguins,                    # the data
       aes(x = flipper_length_mm,   # mapping: column -> x
           y = body_mass_g,         # mapping: column -> y
           color = species)) +      # mapping: column -> colour
  geom_point()                      # geom: draw a dot per row
```

## Same data, different geom

```{r}
#| echo: true
#| fig-width: 7
#| fig-height: 3.8
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_boxplot()
```

A **box** summarises **many** rows. The mappings still mean the same thing.

## Mapping vs. setting

::: {.columns}
::: {.column width="50%"}
**Inside `aes()` = mapping** (from data)
```r
geom_point(aes(color = species))
```
:::
::: {.column width="50%"}
**Outside `aes()` = setting** (one fixed value)
```r
geom_point(color = "navy")
```
:::
:::

The most common beginner mix-up. Colour from a **column** vs. one **fixed** colour.

## The pipe: `|>`

`|>` means **"take this data, then…"**

```{r}
#| echo: true
#| fig-width: 6
#| fig-height: 3.4
penguins |>
  filter(species == "Gentoo") |>
  ggplot(aes(flipper_length_mm, body_mass_g)) +
  geom_point()
```

## Quick check — before you build {.your-turn}

{{< include _wooclap.qmd >}}

## Your turn — Exercise A {.your-turn}

Open `notebooks/01-visualization.qmd`, **Part A**. For each plot, ask:

1. What is the **mark**?
2. Which **column** → which **property**?
3. How many **rows** make one mark?

Then build a scatter one line at a time, and swap in a boxplot.

# Distributions {background-color="#1c4966"}

### What does **one** variable look like?

## One variable, many values

So far every plot compared **two** things (x vs y).

Sometimes you have just **one** column of numbers — say, the body mass of every
penguin — and you want to see its **shape**:

- Where do most values sit?
- Are they spread out or bunched up?
- Are there any surprises?

## `geom_histogram` — count rows into bins

```{r}
#| echo: true
#| fig-width: 7
#| fig-height: 3.6
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(binwidth = 250)
```

Cut the range into **bins**, then count how many **rows** fall in each. The bar's
height *is* a count of rows. (Still: every row lands in exactly one mark.)

## The boxplot is a distribution too

```{r}
#| echo: true
#| fig-width: 7
#| fig-height: 3.4
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

A histogram shows one distribution in detail. A **boxplot** squeezes each
distribution into one **box** — handy for comparing groups side by side.

## Same question, different geoms

- **`geom_histogram`** — the full shape of *one* distribution
- **`geom_boxplot`** — a compact summary, great for *comparing* groups
- Both answer: *"what do these numbers look like?"*

## Your turn — Exercise B {.your-turn}

In `notebooks/01-visualization.qmd`, **Part B**:

1. Make a **histogram** of `flipper_length_mm`.
2. Change the `binwidth` and watch the shape change.
3. Make a **boxplot** of `body_mass_g` by `species`.

## Quick check — reshaping data {.your-turn}

{{< include _wooclap.qmd >}}

# Tidy data {background-color="#1c4966"}

### Real data rarely arrives plot-ready

## Data off a machine

A plate reader gives you one row per sample, with a **separate column for each
gene** it measured:

```{r}
#| echo: false
plate |> head(4) |> knitr::kable()
```

Four measurements are crammed into **one row**. This is *wide* format — easy for
the machine, hard for ggplot.

## ggplot wants **tidy** data

The rule: **one row = one observation**, **one column = one variable**.

One measurement of one gene in one sample *is* an observation — so it should be
its **own row**, with a `gene` column and a `signal` column.

## `pivot_longer` — from wide to long

```{r}
#| echo: true
plate_long <- plate |>
  pivot_longer(gene_A:gene_D, names_to = "gene", values_to = "signal")
```

```{r}
#| echo: false
plate_long |> head(4) |> knitr::kable()
```

The gene **column names** become values in a `gene` column; the numbers stack
into a `signal` column. Now each row is one measurement.

## The payoff: now you can plot it

```{r}
#| echo: true
#| fig-width: 7
#| fig-height: 3.4
ggplot(plate_long, aes(x = gene, y = signal, fill = group)) +
  geom_boxplot()
```

Wide data couldn't make this plot. One `pivot_longer` unlocked it — and the
treatment story (genes C & D switch on) jumps right out.

## Your turn — Exercise C {.your-turn}

In `notebooks/01-visualization.qmd`, **Part C**:

1. `read_csv()` the wide plate file and look at it.
2. `pivot_longer()` the gene columns into `gene` and `signal`.
3. Plot `signal` by `gene`, filled by `group`.

## What you learned today

- Every **row** becomes a **mark**; the **geom** draws it; `aes()` **maps**
  columns to properties.
- **Distributions:** `geom_histogram` and `geom_boxplot` show the shape of numbers.
- **Tidy data:** real data often arrives *wide*; `pivot_longer` makes it plottable.
